A Monte Carlo Synthetic-Acceleration Method for Solving the Thermal Radiation Diffusion Equation

### 2012-08-10 doc/jcomp2012/jcomp2012.tex
updates to jcomp draft

    We present a synthetic-acceleration based Monte Carlo method for
    solving the 1T thermal radiation diffusion equation in 3
    dimensions. We show that this method can be an effective solver
    for sparse matrix systems and is competitive with deterministic
    methods including preconditioned Conjugate Gradient. We also
    discuss various aspects of preconditioning the method and its
    general applicability to broader classes of problems.

@@

In Ref.~\cite{evans_2003}, we developed a residual Monte Carlo method
that was an application of Halton's Sequential Monte Carlo
method~\cite{halton_1962,halton_1994} to the 1D equilibrium (1T)
thermal radiation diffusion equation.  At that time, Halton's method
was not widely known in computational physics.  While extending our
method to 3D, we have discovered that the Sequential Monte Carlo
method is actually a variant of iterative refinement cast as a
residual method.  Using Monte Carlo to accelerate a fixed-point
(Richardson) iteration, our new method surpasses our previous 1D
method and allows for extension to more general solution techniques.

@@

In \S~\ref{sec:monte-carlo-matrix} we review Monte Carlo methods for
solving matrix systems. In \S~\ref{sec:iter-refin-monte} we discuss
our new synthetic-acceleration scheme and connect with previous work
by Halton~\cite{halton_1994} and others~\cite{evans_2003}.  The 1T
radiation diffusion equation and its discrete form will be derived in
\S~\ref{sec:discr-form-radi}.  We discuss using Monte Carlo
Synthetic-Accleration to solve the discrete radiation diffusion
equation in \S~\ref{sec:solut-radi-diff}.  Results from two 3D test
problems are given in \S~\ref{sec:results}.  We finish with
conclusions and ideas for future work in \S~\ref{sec:conclusions}.

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Matrix Solution Methods}
\label{sec:monte-carlo-matrix}

Consider the following matrix equation:

\begin{equation}
  \vA\vx = \vb\:,
  \label{eq:Ax=b}
\end{equation}

which can be written,

\begin{equation}
  \begin{split}
    \vx &= (\vI - \vA)\vx + \vb\\
    &= \vH\vx + \vb\:.
  \end{split}
  \label{eq:point_iteration}
\end{equation}

Here, $\vH = (\vI - \vA)$ is the \latin{iteration matrix}.  When the
spectral radius, $\rho(\vH)$, of the iteration matrix satisfies

\begin{equation}
  \rho(\vH) < 1\:,
\end{equation}

we can expand $\vA^{-1}$ using the \latin{Neumann Series},

\begin{equation}
  \vA^{-1} = (\vI - \vH)^{-1} =  \sum_{k=0}^{\infty} \vH^k\:.
\end{equation}

Thus, when $\rho(\vH) < 1$ we can recast the solution vector $\vx$ as
a series,

\begin{equation}
  \begin{split}
    \vx &= (\vI - \vH)^{-1}\vb\\
    &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb + \ldots\:.
    \label{eq:neumann_series}
  \end{split}
\end{equation}

With this knowledge in hand, we can write an iterative method that
solves Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration),

\begin{equation}
  \vx^{k+1} = \vH\vx^k + \vb\:.
  \label{eq:richardson}
\end{equation}

Equation~(\ref{eq:richardson}) will converge when $\rho(\vH) < 1\ 
\forall\ \vb \in \mathcal{R}^{N}$~\cite{kelley_1995}.

\subsection{Direct Method}
\label{sec:direct-method}

Now we consider a Monte Carlo method that can be used to estimate the
solution to Eq.~(\ref{eq:Ax=b})~\cite{hammersley_1964}.  First,
rewrite Eq.~(\ref{eq:neumann_series}) in order to calculate a
component of $\vx$,

\begin{equation}
  \begin{split}
    x_i &= (\vb)_i + (\vH\vb)_i + (\vH^2\vb)_i + (\vH^3\vb)_i +
    \ldots\\
    &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:.
  \end{split}
  \label{eq:neumann_series_i}
\end{equation}

The terms in the Neumann series can be interpreted as a series of
transitions from $i_{m-1}\rightarrow i_m$ that can be simulated by a
random walk.  Let $X$ be a random variable sampled from a random walk
with $k$ events (particles in the context of radiative transfer) that
initiates in state $i$,

\begin{equation}
  \begin{split}
    X(i_0 = i) &= \sum_{m=0}^{k}W_m b_{i_m}\\
    &= \sum_{m=0}^{k}w_{i,i_1}w_{i_1,i_2}\ldots
    w_{i_{m-1},i_m}b_{i_m}\:.
  \end{split}
  \label{eq:estimator_xi}
\end{equation}

Here, the particle weight on the $m^\text{th}$ step is denoted $W_m$,
and each particle starts with unit weight. At every step,
contributions to each component of $X$ are generated only in the
starting state of the particle, $i_0$, by the source term in the
current state, $b_{i_m}$. Therefore, for a given random walk
permutation, each state must be used as a starting value in order to
calculate its component of the expectation value (i.e. a state of size
$N$ will have $N$ particles per random walk permutation). When the
particle transitions between states, for example $i\rightarrow j$, the
particle weight is multiplied by the transition factor.

\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:,
  \label{eq:weight}
\end{equation}

where $p_{ij}$ denotes the probability of transitioning from state $i$
to $j$. Then, the expected value of $X$ is

\begin{equation}
  \begin{split}
    E[X(i_0=i)] &= \sum_{\nu}P_\nu X_\nu\\
    &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\dots w_{i_{k-1},i_k}b_{i_k}\\
    &= x_i\:,
  \end{split}
  \label{eq:expectation_xi}
\end{equation}

where $\nu$ denotes a particular random walk permutation.
Therefore, the estimator in Eq.~(\ref{eq:estimator_xi}) is an unbiased
estimator of the components of $\vx$ provided $\rho(\vH) < 1$.

We are left to define the transition probabilities. The most
straightforward approach is to set

\begin{equation}
  p_{ij} = \frac{|h_{ij}|}{\sum_{j}|h_{ij}|}\:.
  \label{eq:probability}
\end{equation}

Each row of the transition probability matrix, $\vP$, represents a
discrete probability density function that can be sampled to select a
new state $j$, given that the current state is $i$.  Random walks can
be terminated in two ways: the matrix can be augmented with a
terminating event equation that describes the probability that a
particle ends its walk, or the random walk can be terminated by weight
cutoff, $W_c$.  Generally, we choose to terminate random walks using a
weight cutoff as opposed to augmenting the matrix such that the random
walk terminates when $W_m < W_c$.

\subsection{Adjoint Method}
\label{sec:adjoint-method}

An alternative approach to the one just described is to calculate
contributions to every component of $\vx$ during the random walk.  In
the mathematical literature this is referred to as an adjoint method;
however, to transport practitioners this approach more closely
resembles a traditional forward method.  In this method, the weight
change from state $i\rightarrow j$ is

\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint-weight}
\end{equation}

The transition probabilities may be calculated as

\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_{j}|h_{ji}|}\:.
  \label{eq:adjoint-probability}
\end{equation}

Note that the indices are reversed so that the probabilities are
normalized over a column, as opposed to the forward method in which
the probabilities are normalized over a row.  This is equivalent to
forming the Neumann series in reverse order.  Correspondingly, the
estimator for this method is

\begin{equation}
  \begin{split}
    X &= \sum_{m=0}^{k}W_m\delta_{i_m,i}\\
    &= \sum_{m=0}^{k}\hat{b}_{i_0}w_{i_0,i_1}w_{i_1,i_2}\ldots
                                  w_{i_{m-1},i_m}\delta_{i_m,i}\:.
  \end{split}
  \label{eq:adjoint-tally}
\end{equation}

Here, $\hat{b}_{i_0}$ is the sampled source and initial particle
weight in state $i_0$.  The Kronecker delta implies that tallies are
only made in the state where the random walk currently resides.  This
is the common approach in standard Monte Carlo transport
simulations. Unlike the direct method, because tallies are made in the
state in which the particle currently resides, we are no longer
required to start a particle in each state for each random walk
permutation. Instead, sampling the source is suffucient and requires
only one particle per permutation.

Similar to the direct method, the adjoint method random walk process
requires a terminating condition.  In all of the work that follows we
utilize a relative weight cutoff.  The weight cutoff is defined as a
fraction of the starting weight such that

\begin{equation}
  W_f = W_c\hat{b}_{i_0}\:,
  \label{eq:weight_cutoff}
\end{equation}

where $W_f$ is the terminating weight of the random walk and $W_c$ is
the input relative weight cutoff. The random walk terminates on the
$m^\text{th}$ step if $W_m < W_f$.

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Synthetic-Acceleration}
\label{sec:iter-refin-monte}

The methods presented in \S~\ref{sec:monte-carlo-matrix} are
characterized by slow convergence rates bound by the Central Limit
Theorem.  Halton~\cite{halton_1962,halton_1994} proposed a staged
residual scheme called Sequential Monte Carlo to speed up the
convergence of these methods.  A variant of this scheme has been
successfully applied to the 1D nonlinear thermal radiation diffusion
equation in Ref.~\cite{evans_2003}.

Concisely, the Sequential Monte Carlo method solves
Eq.~(\ref{eq:Ax=b}) using the adjoint solution technique described in
\S~\ref{sec:adjoint-method}.  The following iteration scheme is
applied:

\begin{subequations}
  \begin{gather}
    \vr^{l+1} = \vb - \vA\vx^{l}\:, \\
    \vA\delta\vx^{l+1} = \vr^{l+1}\:, \label{eq:residual_solve}\\
    \vx^{l+1} = \vx^{l} + \delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}

In this scheme, the Monte Carlo adjoint method is used to estimate the
solution to Eq.~(\ref{eq:residual_solve}), and the residual is
iterated to convergence.  This iteration sequence is closely related
to iterative refinement; the exception being that there is no update
to to $x^{l+1}$ between residual iterations.

We propose a modification of this scheme that uses Monte Carlo as a
synthetic-acceleration for the fixed-point iteration given by
Eq.~(\ref{eq:richardson}). We begin by subtracting
Eq.~(\ref{eq:richardson}) from Eq:~(\ref{eq:point_iteration}):

\begin{equation}
  \delta \vx^{l+1} = (\vI - \vA)\delta \vx^{l}\:,
  \label{eq:synthetic_error}
\end{equation}

\begin{equation}
  \delta \vx^{l+1} = \vx - \vx^{l+1}
  \label{eq:lplus1error}
\end{equation}

is defined as the error at iteration $l+1$. We then subtract
$(\vI-\vA) \delta \vx^{l+1}$ from Eq.~(\ref{eq:synthetic_error}),
giving a linear system to solve for the error:

\begin{equation}
  \begin{split}
    \vA \delta \vx^{l+1} &= (\vI - \vA)(\vx^{l+1} - \vx^{l}) \\
    &= \vr^{l+1}\:.
  \end{split}
  \label{eq:synthetic_error2}
\end{equation}

The following scheme will then converge in one iteration:

\begin{subequations}
  \begin{gather}
    \vx^{l+1} = (\vI - \vA)\vx^l + \vb\:,\\
    %%
    \vA \delta \vx^{l+1} = \vr^{l+1}
    \label{eq:synthetic-residual-solve}\\ 
    %% 
    \vx=\vx^{l+1}+\delta\vx^{l+1}\:.
  \end{gather}
  \label{eq:general-synthetic}
\end{subequations}

To accelerate the fixed-point method and avoid inverting $\vA$
directly in Eq.~(\ref{eq:synthetic-residual-solve}), we can instead
use the above ideas to create a new iterative scheme by using the
Monte Carlo method to estimate the solution error. With this in hand,
the Fixed-Point Monte Carlo Synthetic-Acceleration (MCSA) method can
be defined as follows:

\begin{subequations}
  \begin{gather}
    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\
    %%
    \vr^{l+1/2} = \vb - \vA\vx^{l+1/2}\:,
    \label{eq:MCSA-residual}\\     
    %%
    \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
    \label{eq:MCSA-residual_solve}\\ 
    %% 
    \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
  \end{gather}
  \label{eq:general-MCSA}
\end{subequations}

The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that
the Monte Carlo solution only approximately inverts this operator.
However, $\hat{\vA}$ is the same operator as $\vA$;
$\hat{\vA}\equiv\vA$.  Thus, we have defined a scheme in which the
initial estimate of $\vx$ in each iteration is updated using a single
fixed-point iteration\footnote{Although, a Krylov method could be used
  instead of a stationary method here.}.  The residual is calculated
and is used as a source to estimate the error, $\delta\vx^{l+1/2}$, by
solving Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
method. The error is then used to calculate an updated iterate of the
solution vector, $\vx^{l+1}$.  The entire sequence is iterated to
convergence based on the following stopping
criterion~\cite{kelley_1995},
\begin{equation}
  \|\vr\|_{\infty} < \epsilon\cdot\|\vb\|_\infty\:.
  \label{eq:stopping-criteria}
\end{equation}

@@

The MCSA method defined in Eq.(~\ref{eq:general-MCSA}) uses the
adjoint method in Eq.~(\ref{eq:MCSA-residual_solve}) to estimate the
error instead of the direct method outlined in
\S~\ref{sec:direct-method}. To demonstrate the effectiveness of the
adjoint method over the direct method within the context of MCSA, we
choose the two-dimensional time-dependent Poisson equation as a simple
model problem:

@@

\begin{equation}
  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
    u_{i,j+1} - 4 u_{i,j}]
  \label{eq:five_point_stencil}
\end{equation}

and a fourth-order nine point stencil both assuming a grid size of
$\Delta$ in both the $i$ and $j$ directions.

\begin{multline}
  \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\
    + u_{i-1,j+1} + u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]
  \label{eq:nine_point_stencil}
\end{multline}

For a single time step solution, we then have the following sparse
linear system to be solved with the MCSA method.

\begin{equation}
  \vA \vu^{n+1} = \vu^n
  \label{eq:poisson_eq_lin_sys}
\end{equation}

@@

A timing study and convergence study is used to demonstrate the
effectiveness of the adjoint method as compared to the direct
method. To assess both the CPU time and number of iterations required
to converge to a solution, a problem of constant $\Delta$ and was used
with varying values of grid size, fixing the spectral radius of the
system at a constant value for each variation. Both the five point and
nine point stencils were used with both the direct and adjoint
solvers. For each case, 50 random walk permuations were simulated per
iteration with a weight cutoff of $1.0E-4$ for the stochastic linear
solver and a convergence tolerance of $1.0E-8$ for the MCSA iterative
solver. All computations presented in this section were completed on a
2.66 GHz Intel Core i7 machine with 8 GB 1067 MHz DDR3 memory
. Fig.~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
case to converge in seconds and Fig.~\ref{fig:poisson_iterations}
gives the number of iterations needed for each case to converge
relative to the problem size.

@@

where we have dropped the subscript $\mathrm{R}$ from $\sigma$ for
clarity.  Solving this system for $\phi_{l+1/2}$ and plugging into
either $F_{l+1/2}^{+}$ or $F_{l+1/2}^{-}$ and setting the resulting
expression equal to Eq.~(\ref{eq:F_l+1/2}), one finds the
face-centered diffusion coefficient,

@@

Solving for $\phi_{1/2}$ and plugging into
Eq.~(\ref{eq:low-boundary-flux}) yields an expression for the low-side
boundary flux,

@@

For clarity, we rewrite the iteration scheme in
Eqs.~(\ref{eq:general-MCSA}) using the diffusion operator notation
from Eq.~(\ref{eq:preconditioned-diffusion}),

>>>>

    We present a synthetic-acceleration based Monte Carlo method for
    solving the 1T thermal radiation diffusion equation in three
    dimensions. The algorithm performance is compared against
    traditional solution schemes using a Marshak benchmark problem and
    a more sophisticated multiple material problem. Our results show
    that not only can our Monte Carlo method be an effective
    solver for sparse matrix systems, but also that it is competitive
    with deterministic methods including preconditioned Conjugate
    Gradient while producing numerically identical results. We also
    discuss various aspects of preconditioning the method and its
    general applicability to broader classes of problems.

@@

In previous work~\cite{evans_2003}, we developed a residual Monte
Carlo method that was an application of Halton's Sequential Monte
Carlo method~\cite{halton_1962,halton_1994} to the one-dimensional
equilibrium (1T) thermal radiation diffusion equation.  At that time,
Halton's method was not widely known in computational physics.  While
extending our method to three dimensions, we have discovered that the Sequential
Monte Carlo method is actually a variant of iterative refinement cast
as a residual method.  Using Monte Carlo to accelerate a fixed-point
(Richardson) iteration, our new method surpasses our previous 1D
method and allows for extension to more general solution techniques.

@@

In \S~\ref{sec:monte-carlo-matrix} we review Monte Carlo methods for
solving matrix systems. In \S~\ref{sec:iter-refin-monte} we discuss
our new synthetic-acceleration scheme and connect it with previous
work by Halton~\cite{halton_1994} and others~\cite{evans_2003}.  The
1T radiation diffusion equation and its discrete form will be derived
in \S~\ref{sec:discr-form-radi}.  We discuss using Monte Carlo
Synthetic-Accleration to solve the discrete radiation diffusion
equation in \S~\ref{sec:solut-radi-diff}.  Results from two 3D test
problems are given in \S~\ref{sec:results}.  We finish with
conclusions and ideas for future work in \S~\ref{sec:conclusions}.

%%---------------------------------------------------------------------------%%
\section{Mathematical Methods}
\label{sec:monte-carlo-matrix}

The idea of using Monte Carlo methods (random walks) to invert linear
systems is not new.  The earliest referenceable work dates to a 1950
paper by Forsythe and Leibler \cite{forsythe}.  Their paper credits
the idea to unpublished work by J.v. Neumann and S.M. Ulam dating back
to the 1940's.  The basic principles of Monte Carlo matrix inversion
were further elucidated in Hammersley and Handscomb's 1964 text
\cite{hammersley_1964}.  These early methods are distinguished by very
slow, statistically noisy convergence properties; thus, they have not
made any significant impact in the physics community.

In order to address the convergence issues plaguing Monte Carlo
solvers, Halton \cite{halton_1962,halton_1994} proposed a Sequential
Monte Carlo method.  This algorithm demonstrated dramatically improved
convergence over regular Monte Carlo.  Nonetheless, this method has
not gained widespread use in computational physics.  In the next
sections, we will briefly describe the Monte Carlo methods that can be
used to solve linear systems and expand them to present our new
synthetic acceleration method.

%%---------------------------------------------------------------------------%%
\subsection{Monte Carlo Matrix Solution Methods}
\label{sec:background}
To establish the mathematical framework for Monte Carlo linear
solvers, we consider the following matrix equation:
\begin{equation}
  \vA\vx = \vb\:,
  \label{eq:Ax=b}
\end{equation}
which can be written,
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vA)\vx + \vb\\ &= \vH\vx + \vb\:.
  \end{split}
  \label{eq:point_iteration}
\end{equation}
When the spectral radius of the iteration matrix ($\vH$) is less than
1, we can expand $\vA^{-1}$ using the \latin{Neumann Series},
\begin{equation}
  \vA^{-1} = (\vI - \vH)^{-1} = \sum_{k=0}^{\infty} \vH^k\:.
\end{equation} 
Thus, when $\rho(\vH) < 1$ we can recast the solution vector $\vx$ as
a series,
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vH)^{-1}\vb\\ &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb
    + \ldots\:.
    \label{eq:neumann_series}
  \end{split}
\end{equation}
With this knowledge in hand, we can write an iterative method that
solves Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration),
\begin{equation}
  \vx^{k+1} = \vH\vx^k + \vb\:.
  \label{eq:richardson}
\end{equation}
Equation~(\ref{eq:richardson}) will converge for all $\vb \in
\mathcal{R}^{N}$ when $\rho(\vH) < 1$~\cite{kelley_1995}.  All of the
Monte Carlo methods that are described in \S\S~\ref{sec:direct-method}
through \ref{sec:iter-refin-monte} rely on estimating the terms in
Eq.~(\ref{eq:neumann_series}) through random walks.

\subsubsection{Direct Method}
\label{sec:direct-method}

Now we consider a Monte Carlo method that can be used to estimate the
solution to Eq.~(\ref{eq:Ax=b})~\cite{hammersley_1964}.  First,
rewrite Eq.~(\ref{eq:neumann_series}) in order to calculate a
component of $\vx$,
\begin{equation}
  \begin{split}
    x_i &= (\vb)_i + (\vH\vb)_i + (\vH^2\vb)_i + (\vH^3\vb)_i +
    \ldots\\ &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:.
  \end{split}
  \label{eq:neumann_series_i}
\end{equation}
The terms in the Neumann series can be interpreted as a series of
transitions from $i_{m-1}\rightarrow i_m$ that can be simulated by a
random walk.  Let $X$ be a random variable sampled from a random walk
with $k$ events that initiates in state $i$,
\begin{equation}
  \begin{split}
    X(i_0 = i) &= \sum_{m=0}^{k}W_m b_{i_m}\\ &=
    \sum_{m=0}^{k}w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{m-1},i_m}b_{i_m}\:.
  \end{split}
  \label{eq:estimator_xi}
\end{equation}
Here, the weight on the $m^\text{th}$ step is denoted $W_m$, and each
random walk starts with unit weight. At every step, contributions to
each component of $X$ are generated only in the starting state of the
history, $i_0$, by the source term in the current state,
$b_{i_m}$. Therefore, for a given random walk permutation, each state
must be used as a starting value in order to calculate its component
of the expectation value (i.e. a state of size $N$ will have $N$
histories per random walk permutation). When there is a transition
between states, for example $i\rightarrow j$, the weight is multiplied
by the transition factor.
\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:,
  \label{eq:weight}
\end{equation}
where $p_{ij}$ denotes the probability of transitioning from state $i$
to $j$. Then, the expected value of $X$ is
\begin{equation}
  \begin{split}
    E[X(i_0=i)] &= \sum_{\nu}P_\nu X_\nu\\ &=
    \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\dots w_{i_{k-1},i_k}b_{i_k}\\ &= x_i\:,
  \end{split}
  \label{eq:expectation_xi}
\end{equation}
where $\nu$ denotes a particular random walk permutation.  Therefore,
the estimator in Eq.~(\ref{eq:estimator_xi}) is an unbiased estimator
of the components of $\vx$ provided $\rho(\vH) < 1$.

We are left to define the transition probabilities. The most
straightforward approach is to set
\begin{equation}
  p_{ij} = \frac{|h_{ij}|}{\sum_{j}|h_{ij}|}\:.
  \label{eq:probability}
\end{equation}
Each row of the transition probability matrix, $\vP$, represents a
discrete probability density function that can be sampled to select a
new state $j$, given that the current state is $i$.  Random walks can
be terminated in two ways: the matrix can be augmented with a
terminating event equation that describes the probability that a
history ends its walk, or the random walk can be terminated by weight
cutoff, $W_c$.  Generally, we choose to terminate random walks using a
weight cutoff as opposed to augmenting the matrix such that the random
walk terminates when $W_m < W_c$.

\subsubsection{Adjoint Method}
\label{sec:adjoint-method}

An alternative approach to the one just described is to calculate
contributions to every component of $\vx$ during the random walk. In
this method, the weight change from state $i\rightarrow j$ is
\begin{equation}
  w_{ij} = \frac{h_{ji}}{p_{ij}}\:.
  \label{eq:adjoint-weight}
\end{equation}
The transition probabilities may be calculated as
\begin{equation}
  p_{ij} = \frac{|h_{ji}|}{\sum_{j}|h_{ji}|}\:.
  \label{eq:adjoint-probability}
\end{equation}
Note that the indices are reversed so that the probabilities are
normalized over a column, as opposed to the forward method in which
the probabilities are normalized over a row.  This is equivalent to
forming the Neumann series in reverse order.  Correspondingly, the
estimator for this method is
\begin{equation}
  \begin{split}
    X &= \sum_{m=0}^{k}W_m\delta_{i_m,i}\\ &=
    \sum_{m=0}^{k}\hat{b}_{i_0}w_{i_0,i_1}w_{i_1,i_2}\ldots
    w_{i_{m-1},i_m}\delta_{i_m,i}\:.
  \end{split}
  \label{eq:adjoint-tally}
\end{equation}
Here, $\hat{b}_{i_0}$ is the sampled source and initial weight in
state $i_0$.  The Kronecker delta implies that tallies are only made
in the state where the random walk currently resides.  This is the
common approach in standard Monte Carlo transport simulations. Unlike
the direct method, because tallies are made in the state in which the
random walk currently resides, we are no longer required to start a
history in each state for each random walk permutation. Instead,
sampling the source is sufficient and requires only one random walk
per permutation.

Similar to the direct method, the adjoint method random walk process
requires a terminating condition.  In all of the work that follows we
utilize a relative weight cutoff.  The weight cutoff is defined as a
fraction of the starting weight such that
\begin{equation}
  W_f = W_c\hat{b}_{i_0}\:,
  \label{eq:weight_cutoff}
\end{equation}
where $W_f$ is the terminating weight of the random walk and $W_c$ is
the input relative weight cutoff. The random walk terminates on the
$m^\text{th}$ step if $W_m < W_f$.

\subsection{Monte Carlo Synthetic-Acceleration}
\label{sec:iter-refin-monte}

The methods presented in \S~\ref{sec:background} are characterized by
slow convergence rates bound by the Central Limit Theorem.
Halton~\cite{halton_1962,halton_1994} proposed a staged residual
scheme called Sequential Monte Carlo to speed up the convergence of
these methods.  A variant of this scheme has been successfully applied
to the 1D non-linear thermal radiation diffusion equation in
Ref.~\cite{evans_2003}.

Concisely, the Sequential Monte Carlo method solves
Eq.~(\ref{eq:Ax=b}) using the adjoint solution technique described in
\S~\ref{sec:adjoint-method}.  The following iteration scheme is
applied:
\begin{subequations}
  \begin{gather}
    \vr^{l} = \vb - \vA\vx^{l}\:, \\ \vA\delta\vx^{l+1} =
    \vr^{l}\:, \label{eq:residual_solve}\\ \vx^{l+1} = \vx^{l} +
    \delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
In this scheme, the Monte Carlo adjoint method is used to estimate the
solution to Eq.~(\ref{eq:residual_solve}), and the residual is
iterated to convergence.  This iteration sequence is closely related
to iterative refinement; the exception being that there is no update
to $x^{l+1}$ between residual iterations.

We propose a modification of this scheme that uses Monte Carlo as a
synthetic-acceleration for the fixed-point iteration given by
Eq.~(\ref{eq:richardson}). We begin by subtracting
Eq.~(\ref{eq:richardson}) from Eq:~(\ref{eq:point_iteration}):
\begin{equation}
  \delta \vx^{l+1} = (\vI - \vA)\delta \vx^{l}\:,
  \label{eq:synthetic_error}
\end{equation}
where
\begin{equation}
  \delta \vx^{l} = \vx - \vx^{l}
\end{equation}
is defined as the error at iteration $l$. We then subtract $(\vI-\vA)
\delta \vx^{l+1}$ from Eq.~(\ref{eq:synthetic_error}), giving a linear
system to solve for the error:
\begin{equation}
  \begin{split}
    \vA \delta \vx^{l+1} &= (\vI - \vA)(\vx^{l+1} - \vx^{l}) \\ &=
    \vr^{l+1}\:.
  \end{split}
  \label{eq:synthetic_error2}
\end{equation}

The following scheme will then converge in one iteration:
\begin{subequations}
  \begin{gather}
    \vx^{l+1} = (\vI - \vA)\vx^l + \vb\:,\\ \vA \delta \vx^{l+1} =
    \vr^{l+1}\;,
    \label{eq:synthetic-residual-solve}\\ 
    \vx=\vx^{l+1}+\delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
To accelerate the fixed-point method and avoid inverting $\vA$
directly in Eq.~(\ref{eq:synthetic-residual-solve}), we can instead
use the above ideas to create a new iterative scheme by using the
Monte Carlo method to estimate the solution error. With this in hand,
the Fixed-Point Monte Carlo Synthetic-Acceleration (MCSA) method can
be defined as follows:
\begin{subequations}
  \label{eq:MCSA-iteration}
  \begin{gather}
    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\  \vr^{l+1/2} = \vb -
    \vA\vx^{l+1/2}\:,
    \label{eq:MCSA-residual}\\     
     \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
    \label{eq:MCSA-residual_solve}\\ 
     \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
  \end{gather}
\end{subequations}
The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that
the Monte Carlo solution only approximately inverts this operator.
Thus, we have defined a scheme in which the initial estimate of $\vx$
in each iteration is updated using a single fixed-point
iteration\footnote{Although, a subspace method could be used instead of
  a stationary method here.}.  The residual is calculated and is used
as a source to estimate the error, $\delta\vx^{l+1/2}$, by solving
Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
method. The error is then used to calculate an updated iterate of the
solution vector, $\vx^{l+1}$.  The entire sequence is iterated to
convergence based on the following stopping
criterion~\cite{kelley_1995},
\begin{equation}
  \|\vr\|_{\infty} < \epsilon\cdot\|\vb\|_\infty\:.
  \label{eq:stopping-criteria}
\end{equation}

@@

The MCSA method defined in Eq.(~\ref{eq:MCSA-residual_solve}) uses the
adjoint method to estimate the error instead of the direct method
outlined in \S~\ref{sec:direct-method}. To demonstrate the
effectiveness of the adjoint method over the direct method within the
context of MCSA, we choose the two-dimensional time-dependent Poisson
equation as a simple model problem:

@@

\begin{equation}
  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
    u_{i,j+1} - 4 u_{i,j}]\:,
  \label{eq:five_point_stencil}
\end{equation}

and a fourth-order nine point stencil both assuming a grid size of
$\Delta$ in both the $i$ and $j$ directions

\begin{multline}
  \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\ + u_{i-1,j+1} +
    u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]\:.
  \label{eq:nine_point_stencil}
\end{multline}

For a single time step solution, we then have the following sparse
linear system to be solved with the MCSA method:

\begin{equation}
  \vA \vu^{n+1} = \vu^n\:.
  \label{eq:poisson_eq_lin_sys}
\end{equation}

@@

A timing study and convergence study is used to demonstrate the
effectiveness of the adjoint method as compared to the direct
method. To assess both the CPU time and number of iterations required
to converge to a solution, a problem of constant $\Delta$ was used
with varying values of grid size, fixing the spectral radius of the
system at a constant value for each variation. Both the five point and
nine point stencils were used with both the direct and adjoint
solvers. For each case, 50 random walk permuations were simulated per
iteration with a weight cutoff of $1.0E-4$ for the stochastic linear
solver and a convergence tolerance of $1.0E-8$ for the MCSA iterative
solver. All computations presented in this section were completed on a
2.66 GHz Intel Core i7 machine with 8 GB 1067 MHz DDR3 memory
. Fig.~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
case to converge in seconds and Fig.~\ref{fig:poisson_iterations}
gives the number of iterations needed for each case to converge
relative to the problem size.

@@

where we have dropped the subscript $\mathrm{R}$ from $\sigma$ for
clarity.  Solving this system for $\phi_{l+1/2}$ and applying to
either $F_{l+1/2}^{+}$ or $F_{l+1/2}^{-}$ and setting the resulting
expression equal to Eq.~(\ref{eq:F_l+1/2}), one finds the
face-centered diffusion coefficient,

@@

Solving for $\phi_{1/2}$ and applying to
Eq.~(\ref{eq:low-boundary-flux}) yields an expression for the low-side
boundary flux,

@@

For clarity, we rewrite the iteration scheme in
Eq.~(\ref{eq:MCSA-iteration}) using the diffusion operator notation
from Eq.~(\ref{eq:preconditioned-diffusion}),

\begin{align*}
   \bphi^{l+1/2} = (\vI - \ve{M}^{-1}\ve{D})\bphi^{l} +
  \ve{M}^{-1}\ve{q}\:, \quad & (\text{fixed-point iteration})\\ 
  \vr^{l+1/2} = \ve{M}^{-1}\ve{q} - \ve{M}^{-1}\ve{D}\bphi^{l+1/2}\:,
  \quad & (\text{compute residual})\\ 
  \ve{M}^{-1}\ve{D}\delta\bphi^{l+1/2} = \vr^{l+1/2}\:, \quad&
  (\text{estimate $\delta\phi$ using adjoint Monte Carlo method})\\ 
  \bphi^{l+1} = \bphi^{l+1/2} + \delta\bphi^{l+1/2}\:. \quad&
  (\text{update}\ \bphi)
\end{align*}
